### 2020-03-12 thesis/content/chapter-related-work.tex
minor corrections

The \acf{wl} algorithm~\cite{Weisfeiler1968}\cite{Cai1992} characterizes a graph $G$ by assigning discrete labels $c \in \mathcal{C}$, called \textit{colors}, to vertex $k$-tuples $(v_1, \dots, v_k) \in \mathcal{V}_G^k$, where $k \in \mathbb{N}$ is the freely choosable \textit{\ac{wl}-dimension}.
A mapping $\chi_{G, k}: \mathcal{V}_G^k \to \mathcal{C}$ is called a \textit{$k$-coloring} of $G$.
\begin{defn}
	A coloring $\chi'$ \textit{refines} $\chi$ ($\chi' \preceq \chi$) iff.\ $\forall a, b \in \mathcal{V}_G^k: \chi(a) \neq \chi(b) \rightarrow \chi'(a) \neq \chi'(b)$, i.e.\ $\chi'$ distinguishes at least those tuples that are distinguished by $\chi$.
\end{defn}
\begin{defn}
	Two colorings $\chi$ and $\chi'$ are \textit{equivalent} ($\chi \equiv \chi'$) iff.\ $\chi \preceq \chi' \land \chi' \preceq \chi$, i.e.\ $\chi$ is identical to $\chi'$ up to color substitutions.
\end{defn}
The $k$-dimensional \ac{wl} algorithm ($k$-\acs{wl}) works by iteratively refining $k$-colorings $\chi_{G, k}^{(0)} \succeq \chi_{G, k}^{(1)} \succeq \dots$ of a given graph $G$ until the convergence criterion $\chi_{G, k}^{(i)} \equiv \chi_{G, k}^{(i+1)}$ is satisfied.
We denote the final, maximally refined $k$-\acs{wl} coloring with $\chi^{*}_{G, k}$.
\begin{defn}
	The color distribution $\mathit{dist}_{\chi_{G, k}}: \mathcal{C} \to \mathbb{N}$ of a $k$-coloring $\chi_{G, k}$ counts each color $c \in \mathcal{C}$ in the coloring, i.e.\ $\mathit{dist}_{\chi_{G, k}}(c) \coloneqq \left|\left\{ v \in \mathcal{V}_G^k\, |\, \chi_{G, k}(v) = c \right\}\right|$. % chktex 21
\end{defn}
\begin{defn}\label{defn:related:wl-distinguishable}
	Two graphs $G$ and $H$ are $k$-\acs{wl} \textit{distinguishable} ($G \mathrel{{\not\simeq}_k} H$) iff.\ there exists a color $c \in \mathcal{C}$ s.t.\ $\mathit{dist}_{\chi^{*}_{G, k}}(c) \neq \mathit{dist}_{\chi^{*}_{H, k}}(c)$.
\end{defn}
As we will see, the way in which \ac{wl} colorings are refined is vertex order invariant;
thus any difference in the final coloring of two graphs always implies the non-isomorphism of the colored graphs, i.e.\ $G \mathrel{{\not\simeq}_k} H \implies G \not\simeq H$.
The opposite does however not necessarily hold;
two $k$-\acs{wl} indistinguishable graphs are not always isomorphic, i.e.\ $\exists\, G, H: G \mathrel{{\simeq}_k} H \land G \not\simeq H$. % chktex 21

@@

As briefly mentioned before, the combinatorial graph Laplacian $L_G$ is not the only possible choice of graph Laplacian.
Note that the eigenvalues of $L_G$ grow with the number of edges\footnote{
	This is trivially shown by $\sum_{i=1}^n \lambda_{G,i} = \Tr(t_G) = \Tr(D_G) = |\mathcal{E}_G|$.
}.
Consequently a small graph $G$ will typically have a large spectral distance to a large graph $H$ (assuming ${|\mathcal{V}_G|} \ll {|\mathcal{V}_H|}$) irrespective of their similarity when ignoring the scale difference.
In domains where the absolute size of a graph should not influence its spectral characterization, it can therefore be useful to normalize the spectrum.
One common way to do so is via the so-called \textit{symmetric normalized Laplacian} $L^{\text{sym}}_G$.
\begin{defn}\label{defn:related:laplacian-sym}
	$L^{\text{sym}}_G \coloneqq D^{-\frac{1}{2}} L_G D^{-\frac{1}{2}}$ is the symmetric normalized Laplacian of $G$.
\end{defn}
The eigenvalues $\lambda^{\text{sym}}_{G, i}$  of this Laplacian all lie in the range $[0, 2]$.
\Cref{fig:related:spectral-normalization} illustrates how this makes it possible to compare the structure of graphs across varying vertex counts.
Besides $L^{\text{sym}}_G$ there are also other ways to normalize the spectrum, e.g.\ by using the random walk Laplacian $L^{\text{rw}}_G \coloneqq D^{-1} L_G$, which will however not be covered here (see \citet{Shuman2013}).
\begin{figure}[ht]
	\centering
	\includegraphics[width=\linewidth]{gfx/related-work/spectral-normalization.pdf}
	\caption[Comparison of the unnormalized $L_G$ spectrum and the normalized $L^{\text{sym}}_G$ spectrum.]{
		Comparison of the unnormalized $L_G$ spectrum and the normalized $L^{\text{sym}}_G$ spectrum.
		Two structurally similar graphs of different size are compared.
		The \textcolor{t_red}{large graph} is derived by replacing each vertex of the \textcolor{t_blue}{small 4-cycle graph} by a triangle.
		While their unnormalized spectra have different absolute ranges (\textcolor{t_blue}{$[0, 4]$} vs.\ \textcolor{t_red}{$[0, 12]$}), the histogram on the right shows that their normalized eigenvalue distributions align.
	}\label{fig:related:spectral-normalization}
\end{figure}

@@

\paragraph{Higher dimensional \ac{wl} \acp{gnn}}
The majority of \ac{gnn} methods uses a neighborhood aggregation scheme which limits their discriminative and, more importantly, computational power to that of 1-\acs{wl} (see \cref{defn:related:wl-count-detect}, \cpageref{defn:related:wl-count-detect}).
To go beyond the limits of 1-\acs{wl}, \citet{Morris2019} proposed the first \ac{gcnn} architecture inspired by the higher dimensional $k$-\acs{wl} algorithms.
Similar to the $k$-LWL kernel which we looked at in \cref{sec:related:gcr:kernel}, their so-called $k$-\acs{gnn} uses the $k$-multiset and neighborhood localization optimizations to keep runtime costs feasible.
The $k$-\ac{gnn} convolution is described by
\begin{align}
	Z^{(t)}[s] \coloneqq \sigma\left( {Z^{(t-1)}[s]} W^{(t)} + \smashoperator[lr]{\sum_{v \in s, u \in \Gamma_G(v)}} {Z^{(t-1)}[s \setminus \{ v \} \cup \{ u \}]} W_{\Gamma}^{(t)} \right) \text{.} \label{eq:related:kgnn-layer}
\end{align}
It convolves the feature vectors of $k$-multisets $s \subseteq \mathcal{V}_G$ instead of single vertices $v \in \mathcal{V}_G$.
Apart from this fundamental difference to 1-\acs{wl} bounded \acp{gnn}, $k$-\acp{gnn} additionally use two learnable parameter matrices $W^{(t)}, W_{\Gamma}^{(t)} \in \mathbb{R}^{d^{(t-1)} \times d^{(t)}}$ instead of the single $\Theta^{(t)}$ used in \acp{gcn}.
This separation of parameter matrix effectively fulfills the same purpose as the \ac{mlp} in \ac{gin}; it makes the implicitly defined $k$-\acs{gnn} hash function injective.
This can be seen by realizing that a stack of $k$-\acs{gnn} convolution layers with $W_{\Gamma}^{(t)} = 0$ for all but the first layer simulates a \ac{mlp}.

>>>>

The \acf{wl} algorithm~\cite{Weisfeiler1968}\cite{Cai1992} characterizes a graph $G$ by assigning discrete labels $c \in \mathcal{C}$, called \textit{colors}, to vertex $k$-tuples $(v_1, \dots, v_k) \in \mathcal{V}_G^k$, where $k \in \mathbb{N}$ is the freely choosable \textit{\ac{wl}-dimension}.
A mapping $\chi_{G, k}: \mathcal{V}_G^k \to \mathcal{C}$ is called a \textit{$k$-coloring} of $G$.
\begin{defn}
	A coloring $\chi'$ \textit{refines} $\chi$ ($\chi' \preceq \chi$) iff.\ $\forall a, b \in \mathcal{V}_G^k: \chi(a) \neq \chi(b) \rightarrow \chi'(a) \neq \chi'(b)$, i.e.\ $\chi'$ distinguishes at least those tuples that are distinguished by $\chi$.
\end{defn}
\begin{defn}
	Two colorings $\chi$ and $\chi'$ are \textit{equivalent} ($\chi \equiv \chi'$) iff.\ $\chi \preceq \chi' \land \chi \succeq \chi'$, i.e.\ $\chi$ is identical to $\chi'$ up to color substitutions.
\end{defn}
The $k$-dimensional \ac{wl} algorithm ($k$-\acs{wl}) works by iteratively refining $k$-colorings $\chi_{G, k}^{(0)} \succeq \chi_{G, k}^{(1)} \succeq \dots$ of a given graph $G$ until the convergence criterion $\chi_{G, k}^{(i)} \equiv \chi_{G, k}^{(i+1)}$ is satisfied.
We denote the final, maximally refined $k$-\acs{wl} coloring with $\chi^{*}_{G, k}$.
\begin{defn}
	The color distribution $\mathit{dist}_{\chi_{G, k}}: \mathcal{C} \to \mathbb{N}$ of a $k$-coloring $\chi_{G, k}$ counts each color $c \in \mathcal{C}$ in the coloring, i.e.\ $\mathit{dist}_{\chi_{G, k}}(c) \coloneqq \left|\left\{ v \in \mathcal{V}_G^k\, |\, \chi_{G, k}(v) = c \right\}\right|$. % chktex 21
\end{defn}
\begin{defn}\label{defn:related:wl-distinguishable}
	Two graphs $G$ and $H$ are $k$-\acs{wl} \textit{distinguishable} ($G \mathrel{{\not\simeq}_k} H$) iff.\ there exists a color $c \in \mathcal{C}$ s.t.\ $\mathit{dist}_{\chi^{*}_{G, k}}(c) \neq \mathit{dist}_{\chi^{*}_{H, k}}(c)$.
\end{defn}
As we will see, the way in which \ac{wl} colorings are refined is vertex order invariant;
thus any difference in the final coloring of two graphs always implies the non-isomorphism of the colored graphs, i.e.\ $G \mathrel{{\not\simeq}_k} H \implies G \not\simeq H$.
The opposite does however not necessarily hold;
two $k$-\acs{wl} indistinguishable graphs are not always isomorphic, i.e.\ $\exists\, G, H: G \mathrel{{\simeq}_k} H \land G \not\simeq H$. % chktex 21

@@

As briefly mentioned before, the combinatorial graph Laplacian $L_G$ is not the only possible choice of graph Laplacian.
Note that the eigenvalues of $L_G$ grow with the number of edges\footnote{
	This is trivially shown by $\sum_{i=1}^n \lambda_{G,i} = \Tr(L_G) = \Tr(D_G) = |\mathcal{E}_G|$.
}.
Consequently a small graph $G$ will typically have a large spectral distance to a large graph $H$ (assuming ${|\mathcal{V}_G|} \ll {|\mathcal{V}_H|}$) irrespective of their similarity when ignoring the scale difference.
In domains where the absolute size of a graph should not influence its spectral characterization, it can therefore be useful to normalize the spectrum.
One common way to do so is via the so-called \textit{symmetric normalized Laplacian} $L^{\text{sym}}_G$.
\begin{defn}\label{defn:related:laplacian-sym}
	$L^{\text{sym}}_G \coloneqq D^{-\frac{1}{2}} L_G D^{-\frac{1}{2}}$ is the symmetric normalized Laplacian of $G$.
\end{defn}
The eigenvalues $\lambda^{\text{sym}}_{G, i}$  of this Laplacian all lie in the range $[0, 2]$.
\Cref{fig:related:spectral-normalization} illustrates how this makes it possible to compare the structure of graphs across varying vertex counts.
Besides $L^{\text{sym}}_G$ there are also other ways to normalize the spectrum, e.g.\ by using the random walk Laplacian $L^{\text{rw}}_G \coloneqq D^{-1} L_G$, which will however not be covered here (see \citet{Shuman2013}).
\begin{figure}[ht]
	\centering
	\includegraphics[width=\linewidth]{gfx/related-work/spectral-normalization.pdf}
	\caption[Comparison of the unnormalized $L_G$ spectrum and the normalized $L^{\text{sym}}_G$ spectrum.]{
		Comparison of the unnormalized $L_G$ spectrum and the normalized $L^{\text{sym}}_G$ spectrum.
		Two structurally similar graphs of different size are compared.
		The \textcolor{t_red}{large graph} is derived by replacing each vertex of the \textcolor{t_blue}{small 4-cycle graph} by a triangle.
		While their unnormalized spectra have different absolute ranges (\textcolor{t_blue}{$[0, 4]$} vs.\ \textcolor{t_red}{$[0, 12]$}), the histogram on the right shows that their normalized eigenvalue distributions align.
	}\label{fig:related:spectral-normalization}
\end{figure}

@@

\paragraph{Higher dimensional \ac{wl} \acp{gnn}}
The majority of \ac{gnn} methods uses a neighborhood aggregation scheme which limits their discriminative and, more importantly, computational power to that of 1-\acs{wl} (see \cref{defn:related:wl-count-detect}, \cpageref{defn:related:wl-count-detect}).
To go beyond the limits of 1-\acs{wl}, \citet{Morris2019} proposed the first \ac{gcnn} architecture inspired by the higher dimensional $k$-\acs{wl} algorithms.
Similar to the $k$-LWL kernel which we looked at in \cref{sec:related:gcr:kernel}, their so-called $k$-\acs{gnn} uses the $k$-multiset and neighborhood localization optimizations to keep runtime costs feasible.
The $k$-\ac{gnn} convolution is described by
\begin{align}
	Z^{(t)}[s] \coloneqq \sigma\left( {Z^{(t-1)}[s]} W^{(t)} + \smashoperator[lr]{\sum_{v \in s, u \in \Gamma_G(v)}} {Z^{(t-1)}[s \setminus \{ v \} \cup \{ u \}]} W_{\Gamma}^{(t)} \right) \text{.} \label{eq:related:kgnn-layer}
\end{align}
It convolves the feature vectors of $k$-multisets $s \subseteq \mathcal{V}_G$ instead of single vertices $v \in \mathcal{V}_G$.
Apart from this fundamental difference to 1-\acs{wl} bounded \acp{gnn}, $k$-\acp{gnn} additionally use two learnable parameter matrices $W^{(t)}, W_{\Gamma}^{(t)} \in \mathbb{R}^{d^{(t-1)} \times d^{(t)}}$ instead of the single $\Theta^{(t)}$ used in \acp{gcn}.
This separation of the parameter matrix effectively fulfills the same purpose as the \ac{mlp} in \ac{gin}; it makes the implicitly defined $k$-\acs{gnn} hash function injective.
This can be seen by realizing that a stack of $k$-\acs{gnn} convolution layers with $W_{\Gamma}^{(t)} = 0$ for all but the first layer simulates a \ac{mlp}.
